Student Log-Data from a Randomized Evaluation of Educational Technology: A Causal Case Study

### 2020-02-04 paper.Rnw
edits to sec 2

\subsection{The Role of Hints\footnote{This subsection draws heavily
  on comments from an anonymous reviewer.}}
Students using the Cognitive Tutor Algebra software work algebra
problems on a computer.
Each problem in the software is associated with a set of skills that
the student must master, and a path (or set of paths) to solving the
problem.
As soon as a student errs in working through a problem---that is, departs from one of the
solution paths---the
software shows the student an error message.
This feedback is tailored to the specific mistake, and
designed to guide the student back to a correct solution path.
Similarly, when a student gets stuck, and is unable to determine the
next step in solving the problem, he or she can ask for a hint.
Like the error feedback, these hints are tailored to the specific
algebra skills corresponding to the next step on the solution path.
Thus, hints and error feedback are crucial aspects of the tutor's
pedagogy.

@@

The empirical literature on the value of hints in intelligent tutors
is mixed \citep[see][for a summary]{goldin2012learner}.
\citet{anderson1995cognitive} compared the performance of students
randomly assigned to different versions of the Cognitive Tutor with
different feedback structures, and found that students who were able
to request hints finished the lesson faster than those who were not,
but could not detect an effect on learning.
\citet{singh2011feedback} used a similar design, comparing two
versions of the ASSISTments intelligent tutor, one in which hints and
immediate error messages were available, and one which did not provide
feedback. Students in the feedback condition showed larger gains on a
posttest, though it is unclear whether these gains were due to hints
or error feedback.
On the other hand, \citet{beck2008does} and \citet{goldin2012learner}, did not randomize hint
availability, but instead used statistical models to control for
observed confounding.
Both papers used students' subsequent performance within the tutor to
estimate hint effects.
\citet{beck2008does} found conflicting results using different
analytical methods to control for confounding.
\citet{goldin2012learner} investigated two sources of heterogeneity in
the effect of a hint request: by student and by the ``level'' of the
hint (the first, second, or final hint available for a given problem).
They found that, indeed, the effect of asking for a hint varies
between students, but on average it is negative for the first two
hints requested on a problem, but positive for the final hint.

@@

To choose $c$, we first fit a modified Rasch mixture model
\citep{rasch} to hint request data.
Specifically, we modeled the probability of a hint request in each problem
$p$, worked by student $i$, as
\begin{equation*}\label{eq:rasch}
Pr(h_{ip})=logit^{-1}(\eta_i+\delta_{s[p]})
\end{equation*}
where $h_{ip}$ is the event that student $i$ requests a hint on
problem $p$, $\eta_i$ is a student parameter, $\delta_{s[p]}$ is a
section-level parameter for $s[p]$, the section that problem $p$ was
drawn from, and $logit^{-1}(x)=\{1+exp(-x)\}^{-1}$ is the inverse logit
function.
Typically, the student parameter $\eta_i$ measures student ability;
here, instead, it measures a student's proclivity to request a hint.
The typical Rasch model includes a problem-level ``difficulty'' parameter, but in our
dataset many problems were worked by very few students, so problem
level parameters would be hard to estimate.
Instead, we included the section-level parameter $\delta_{s[p]}$,
which measures the extent to which students tend to request hints on
problems in section $s$.
This essentially assumes that all problems within each section are equally
likely to elicit a hint.

>>>>

\subsection{The Role of Hints\footnote{This subsection draws heavily
  on comments from an anonymous reviewer.}}
Students using the Cognitive Tutor Algebra software work algebra
problems on a computer.
Each problem in the software is associated with a set of skills that
the student must master in order to solve the problem.
The software also specifies a ``solution path'' for each problem---a sequence
of actions (such as ``multiply both sides of the equation by 3'')
students must take in order to solve the problem.
If there is more than one way to solve a problem, the software specifies
multiple correct solution paths students may follow.
As soon as a student errs in working through a problem---that is, departs from one of the
solution paths---the software shows the student an error message.
This feedback is tailored to the specific mistake, and
designed to guide the student back to a correct solution path.
Similarly, when a student gets stuck, and is unable to determine the
next step in solving the problem, he or she can ask for a hint.
Like the error feedback, these hints are tailored to the specific
algebra skills corresponding to the next step on the solution path.
Thus, hints and error feedback are crucial aspects of the tutor's
pedagogy.

@@

The empirical literature on the value of hints in intelligent tutors
is mixed \citep[see][for a summary]{goldin2012learner}.
\citet{anderson1995cognitive} compared the performance of students
randomly assigned to different versions of the Cognitive Tutor with
different feedback structures, and found that students who were able
to request hints finished the lesson faster than those who were not,
but could not detect an effect on learning.
\citet{singh2011feedback} used a similar design, comparing two
versions of the ASSISTments intelligent tutor, one in which hints and
immediate error messages were available, and one which did not provide
feedback. Students in the feedback condition showed larger gains on a
posttest, though it is unclear whether these gains were due to hints
or error feedback.
\citet{beck2008does} and \citet{goldin2012learner}, did not randomize hint
availability, but instead used statistical models to control for
observed confounding.
Both papers used students' subsequent performance within the tutor to
estimate hint effects.
\citet{beck2008does} found conflicting results using different
analytical methods to control for confounding.
\citet{goldin2012learner} investigated two sources of heterogeneity in
the effect of a hint request: by student and by the ``level'' of the
hint (the first, second, or final hint available for a given problem).
They found that the effect of asking for a hint varies
between students, but on average it is negative for the first two
hints requested on a problem, but positive for the final hint.

@@

To choose $c$, we first fit a modified Rasch mixture model
\citep{rasch} to hint request data.
Specifically, we modeled the probability of a hint request in each problem
$p$, worked by student $i$, in which $i$ requested a hint and/or
committed an error, as 
\begin{equation}\label{eq:rasch}
Pr(h_{ip})=logit^{-1}(\eta_i+\delta_{s[p]})
\end{equation} 
where $h_{ip}$ is the event that student $i$ requests a hint on 
problem $p$, $\eta_i$ is a student parameter, $\delta_{s[p]}$ is a
section-level parameter for $s[p]$, the section that problem $p$ was
drawn from, and $logit^{-1}(x)=\{1+exp(-x)\}^{-1}$ is the inverse logit
function.
Typically, the student parameter $\eta_i$ measures student ability;
here, instead, it measures a student's proclivity to request a hint.
The typical Rasch model includes a problem-level ``difficulty'' parameter, but in our
dataset many problems were worked by very few students, so problem
level parameters would be hard to estimate.
Instead, we included the section-level parameter $\delta_{s[p]}$,
which measures the extent to which students tend to request hints on
problems in section $s$.
This essentially assumes that all problems within each section are equally
likely to elicit a hint request.
